---
title: "Browsing Disease and Cancer Information from cBioPortal"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Browsing cBioPortal Disease Information}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output:
  BiocStyle::html_document: 
    number_sections: yes
    toc: yes
    toc_depth: 4
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", 
                      collapse = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

# Introduction

The cBioPortal for Cancer Genomics provides a comprehensive resource for 
exploring cancer genomics data from various studies. The OmicsMLRepoR package 
offers harmonized and ontology-enhanced metadata from cBioPortal, making it 
easier to browse and filter cancer and disease information across multiple 
studies.

This vignette demonstrates how to: 

1. Load harmonized cBioPortal metadata
2. Browse available cancer types and diseases
3. Use ontology-based filtering to find related diseases
4. Explore disease relationships through ontology hierarchies
5. Extract disease-specific subsets for downstream analysis

# Load packages

```{r load_packages}
suppressPackageStartupMessages({
    library(OmicsMLRepoR)
    library(dplyr)
    library(cBioPortalData)
})
```

# Load cBioPortal metadata

The `getMetadata` function downloads the harmonized cBioPortal metadata, which 
includes ontology mappings for various clinical attributes including cancer 
types and diseases.

```{r load_metadata}
cbio <- getMetadata("cBioPortal")
cbio
```

# Explore available disease information

## View metadata structure

Let's first examine what disease-related columns are available in the 
harmonized metadata:

```{r view_columns}
# Display columns containing disease or cancer information
disease_cols <- colnames(cbio)[grep("disease|cancer|oncotree_code", 
                                     colnames(cbio), 
                                     ignore.case = TRUE)]
disease_cols
```

## Explore cancer type oncotree_codes

The `oncotree_code` column contains standardized cancer type abbreviations from 
cBioPortal studies: 

```{r explore_oncotree_codes}
# Get unique cancer types
unique_cancers <- cbio %>%
    dplyr::select(oncotree_code) %>%
    distinct() %>%
    arrange(oncotree_code)

head(unique_cancers, 20)
```

## Check ontology-mapped attributes

Columns with ontology mappings enable robust searching using the `tree_filter` 
function:

```{r ontology_attributes}
# Find columns with ontology term IDs
onto_cols <- colnames(cbio)[grep("_ontology_term_id", colnames(cbio))]
onto_cols

# Get the corresponding attribute names
gsub("_ontology_term_id", "", onto_cols)
```

# Browse diseases using ontology

The `tree_filter` function enables powerful ontology-based searching that 
automatically includes: 

- **Case-insensitive matching**: Search terms work regardless of capitalization
- **Synonyms**: Alternative names for the same disease
- **Descendants**:  All child terms in the ontology hierarchy

## Example 1: Search for breast cancer

```{r breast_cancer}
# Find all samples related to breast cancer
breast_cancer <- cbio %>%
    tree_filter(cancer_type, "breast cancer")

# View unique cancer types found
breast_cancer %>%
    select(cancer_type) %>%
    distinct()

# Count samples
nrow(breast_cancer)
```

## Example 2: Search using cancer type abbreviations

```{r search_abbreviations}
# Search for lung cancer studies
lung_cancer <- cbio %>%
    tree_filter(oncotree_code, c("LUAD", "LUSC"))

# View results
lung_cancer %>%
    select(oncotree_code, studyId) %>%
    distinct() %>%
    arrange(oncotree_code)
```

## Example 3: Find related cancer types using ontology hierarchy

Ontology-based searching automatically includes descendant terms.  For example, 
searching for "carcinoma" will find all specific types of carcinomas:

```{r carcinoma_search}
# Search for carcinomas (includes all descendant types)
carcinomas <- cbio %>%
    tree_filter(cancer_type, "carcinoma")

# See the variety of carcinoma types found
carcinomas %>%
    select(cancer_type) %>%
    distinct() %>%
    head(15)

# Count total samples
nrow(carcinomas)
```

## Example 4: Compare search terms

Demonstrate the power of ontology-based searching by comparing different 
search terms: 

```{r compare_searches}
# Search using different terms for colorectal cancer
crc_oncotree_code <- cbio %>%
    tree_filter(oncotree_code, "CRC") %>%
    nrow()

crc_full <- cbio %>%
    tree_filter(cancer_type, "colorectal cancer") %>%
    nrow()

crc_synonym <- cbio %>%
    tree_filter(cancer_type, "colorectal carcinoma") %>%
    nrow()

# Compare results
data.frame(
    search_term = c("CRC", "colorectal cancer", "colorectal carcinoma"),
    samples_found = c(crc_oncotree_code, crc_full, crc_synonym)
)
```

# Advanced filtering with multiple conditions

## Combine multiple search terms with OR logic

```{r multiple_or}
# Find samples from breast, ovarian, or endometrial cancers
gynecologic_cancers <- cbio %>%
    tree_filter(cancer_type, c("breast cancer", "ovarian cancer", 
                            "endometrial cancer"), 
                logic = "OR")

# Summarize by cancer type
gynecologic_cancers %>%
    group_by(cancer_type) %>%
    summarise(n_samples = n()) %>%
    arrange(desc(n_samples)) %>%
    head(10)
```

## Combine tree_filter with dplyr for complex queries

```{r complex_queries}
# Find lung cancer samples from female patients
lung_female <- cbio %>%
    tree_filter(cancer_type, "lung cancer") %>%
    filter(sex == "female")

# Count results
nrow(lung_female)

# Find melanoma samples with treatment information
melanoma_treated <- cbio %>%
    tree_filter(cancer_type, "melanoma") %>%
    filter(!is.na(treatment_name))

# View treatment types
melanoma_treated %>%
    select(cancer_type, treatment_name, treatment_type) %>%
    distinct() %>%
    head(10)
```

# Explore disease hierarchies

## Find broad disease categories

Search for broad disease terms to capture multiple related cancer types:

```{r broad_categories}
# Find all samples under "neoplasm" (cancer)
all_neoplasms <- cbio %>%
    tree_filter(cancer_type, "neoplasm")

# See the diversity of cancer types
all_neoplasms %>%
    select(cancer_type) %>%
    distinct() %>%
    nrow()
```

## Navigate from general to specific

```{r hierarchy_navigation}
# Start broad:  all lymphomas
all_lymphomas <- cbio %>%
    tree_filter(cancer_type, "lymphoma")

lymphoma_types <- all_lymphomas %>%
    select(cancer_type) %>%
    distinct()

nrow(lymphoma_types)
head(lymphoma_types, 10)

# Get more specific:  Hodgkin lymphoma
hodgkin <- cbio %>%
    tree_filter(cancer_type, "hodgkin lymphoma")

nrow(hodgkin)
```

# Export disease-specific subsets

Once you've identified samples of interest, you can export them for further 
analysis:

```{r export_subset}
# Create a subset for pancreatic cancer
pancreatic_subset <- cbio %>%
    tree_filter(cancer_type, "pancreatic cancer") %>%
    select(curation_id, studyId, cancer_type, sex, age, 
           treatment_name, treatment_type)

# View summary
head(pancreatic_subset)

# Export to CSV (optional)
# write.csv(pancreatic_subset, "pancreatic_cancer_samples.csv", 
#           row.names = FALSE)
```

# Tips for effective searching

## 1. Try multiple search terms

Different studies may use different terminology. Try variations:

```{r search_variations}
# These might return different results depending on study annotation
search_terms <- c("CRC", "colorectal", "colon cancer", "rectal cancer")

results <- sapply(search_terms, function(term) {
    cbio %>%
        tree_filter(cancer_type, term) %>%
        nrow()
})

data.frame(
    term = search_terms,
    samples = results
)
```

## 2. Use ontology IDs for precise searching

If you know the specific ontology ID, you can search using it directly:

```{r ontology_id_search}
# Example: Search using a specific disease ontology ID
# (Replace with actual IDs from your data)
cbio %>%
    select(cancer_type, cancer_type_ontology_term_id) %>%
    distinct() %>%
    head(10)
```

## 3. Check available ontologies

Different attributes may use different ontologies:

```{r check_ontologies}
# View which ontologies are used in the metadata
# Ontology IDs typically start with ontology prefixes (e.g., DOID, NCIT)
cbio %>%
    select(contains("ontology_term_id")) %>%
    slice(1:5)
```

# Comparison with non-harmonized metadata

The harmonized metadata from OmicsMLRepoR offers significant advantages over 
working directly with raw cBioPortal data: 

**Before harmonization (raw cBioPortal metadata):**
- Case-sensitive searches
- Inconsistent naming across studies
- Manual synonym handling required
- Difficult to find related diseases

**After harmonization (OmicsMLRepoR):**
- Case-insensitive searching
- Standardized terminology
- Automatic synonym recognition
- Ontology-based hierarchical searching
- Easy filtering with `tree_filter`

# Session information

```{r session_info}
sessionInfo()
```

# Further reading

For more information about OmicsMLRepoR: 

- See the Quickstart vignette: `vignette("Quickstart", package = "OmicsMLRepoR")`
- Visit the project repository: https://github.com/shbrief/OmicsMLRepoR
- Learn about the harmonization process: https://github.com/waldronlab/OmicsMLRepoData
```